### 2024-07-08 cognitive-bias.qmd
20240707 - heuristics and cognitive biases

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called *heuristics*.
These mental shortcuts can be helpful.
They can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
However, these mental shortcuts can also lead us astray and make systematic errors in our judgments and predictions.
*Cognitive biases* are systematic errors in thinking.

@@

@@

>>>>

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called *heuristics*.
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
However, these mental shortcuts can also lead us astray and make systematic errors in our judgments and predictions.
*Cognitive biases* are systematic errors in thinking.

@@

Three important heuristics used in judgment and prediction in the face of uncertainty include [@Tversky1974]:

- availability
- representativeness
- anchoring and adjustment

The *availability heuristic* refers to the tendency for a person's judgments or predictions about the frequency or probability of something to be made based on how readily instances can be brought to mind.
For instance, when making fantasy predictions about a player, more recent big performance games may more easily come to mind compared to lower-scoring games and games that occurred longer ago.
Thus, a manager may be more inclined to pick players to start who had more recent, stronger performances rather than players who have higher long-term averages.

The *representativess heuristic* refers to the tendency for a person's judgments or predictions about individuals to be made based on how similar the individual is to the person's existing mental prototypes.
For instance, when coming out of college, Tight End Kyle Pitts [drew comparisons](https://247sports.com/article/kyle-pitts-lebron-james-2021-nfl-draft-florida-gators-football-163882176) to the "LeBron James" of Tight Ends (archived at <https://perma.cc/JQB5-XPVL>).
The idea that his athletic profile leads him to be similar to the prototype of LeBron James may have led him to be too highly drafted by fantasy managers in his first seasons.

::: {.content-visible when-format="html:js"}

Here is a video of Kyle Pitts drawing comparisons to the LeBron James of Tight Ends:

![Video of Kyle Pitts drawing comparisons to the LeBron James of Tight Ends. From: <https://x.com/GetUpESPN/status/1380165126108672001> (archived at <https://perma.cc/JW8E-KV2C>)](images/kylePitts.mp4){width=100%}

The representativeness heuristic has been observed in gambling markets for predicting team wins in the National Football League (NFL) [@Woodland2015] and in decision making in fantasy soccer [@Kotrba2020].

The *anchoring and adjustment heuristic* refers to the tendency for a person's judgments or predictions to be made with a reference point—an anchor—as a starting point from which they adjust their estimates upward or downward.
The anchor is often inaccurate and given too much weight in the person's calculation, and too little adjustment is made to the anchor.
For instance, a manager is trying to predict how many fantasy points a top Running Back may score.
The player scored 300 fantasy points last season, but the team added a stronger backup Running Back and changed the Offensive Coordinator to be a more pass-heavy offense.
The manager may use 300 fantasy points as an anchor (based on the player's performance last season), and may adjust downward 15 points to account for the offseason changes.
However, it is possible that this downward adjustment is insufficient to account not only for the offseasons changes but also for potential regression effects.
Regression effects are discussed further below.

@@

Examples of cognitive biases that result from one or more of these heuristics include:

- base rate neglect
- forgetting about regression to the mean
- overconfidence
- confirmation bias
- hindsight bias
- loss aversion bias
- endowment bias
- Dunning–Kruger effect

## Examples of Fallacies {#sec-fallacies}

Fallacies are mistaken beliefs that are often due to heuristics and cognitive biases.
Examples of fallacies include:

- sunk cost fallacy
- hot hand fallacy
- gambler's fallacy

### 2024-07-08 draft.qmd
20240627 - sessionInfo() for reproducibility
20240627 - sessionInfo()
20240707 - heuristics and cognitive biases

There is no one "right" draft strategy.
Sometimes it works best to "zig" when everyone else is "zagging".
For instance, if you notice that everyone else is drafting Wide Receivers, this may mean that other managers are over-valuing Wide Receivers, and this could be a nice opportunity to draft a Running Back for good value.

@@

>>>>

There is no one "right" draft strategy.
As noted by @Lee2022 in their analysis of fantasy drafts, the effectiveness of any draft strategy depends on the strategies of the other managers in the league.
Sometimes it works best to "zig" when everyone else is "zagging".
For instance, if you notice that everyone else is drafting Wide Receivers, this may mean that other managers are over-valuing Wide Receivers, and this could be a nice opportunity to draft a Running Back for good value.

@@

An alternative to the "stars and scrubs" approach is to wait to draft more "high-value" players after other managers have over-paid for players.

::: {.content-visible when-format="html"}

## Session Info {#sec-draftSessionInfo}

```{r}
sessionInfo()
```

:::
